---
title: "k-means and pca"
author: "Anna Khazenzon"
date: "February 13, 2015"
output: html_document
---

#clustering lab, ISLR 10.4 & 10.5.1

## k-means

Simulate data with mean shift between first and second 25 obs.

``` {r}
set.seed(2)
x = matrix(rnorm(50*2), ncol=2)
x[1:25,1] = x[1:25,1] + 3
x[1:25,2] = x[1:25,2] - 4
```

K-means with K = 2

``` {r}
km.out = kmeans(x, 2, nstart = 20)
km.out$cluster
plot(x, col=(km.out$cluster + 2), main="K-Means Clustering results with K=2", xlab="", ylab="", pch=20, cex=2)
```

K-means with K = 3

``` {r}
set.seed(4)
km.out = kmeans(x, 3, nstart = 20)
km.out
plot(x, col=(km.out$cluster + 2), main="K-Means Clustering results with K=3", xlab="", ylab="", pch=20, cex=2)
```

K-means with multiple initial cluster assignments

``` {r}
set.seed(3)
km.out = kmeans(x, 3, nstart = 1)
km.out$tot.withinss
km.out = kmeans(x, 3, nstart = 20)
km.out$tot.withinss
```

Higher nstart minimizes within-cluster sum of squares. Always run K-means with a large value of nstart (eg 20, 50) to avoid local optima.

Setting the seed is important in order to be able to reproduce the initial cluster assignments.

## PCA

Examine data.

``` {r}
states = row.names(USArrests)
states

names(USArrests)

apply(USArrests, 2, mean) # v diff means; 3x as many assaults as rapes
apply(USArrests, 2, var) # v diff variance; not surprising, since nums not comparable; hint: need to scale! otherwise PCA driven by Assault
```

Perform PCA.

``` {r}
pr.out = prcomp(USArrests, scale=T)
pr.out$center # mean of vars used for scaling
pr.out$scale # sd of vars used for scaling
pr.out$rotation # principal component loadings
```

There are min(n-1, p) informative principal components in a data set with n observations and p variables. 

``` {r}
dim(pr.out$x) #cols are principal component score vectors

biplot(pr.out, scale=0)
```

``` {r}
pr.out$sdev #sd of each principal component
pr.var = pr.out$sdev^2
pve = pr.var/sum(pr.var); pve # percent variance explained by each comp

plot(pve, xlab = "Principal Component", ylab = "Proportion of Variance Explained", ylim = c(0,1), type = 'b')

plot(cumsum(pve), xlab = "Principal Component", ylab = "Cumulative Proportion of Variance Explained", ylim = c(0,1), type = 'b')
```






